\documentclass[sigconf]{acmart}
\usepackage{tikz}
\usetikzlibrary{arrows.meta,positioning,fit}
\usepackage{booktabs}

\settopmatter{printacmref=false}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\setcopyright{none}
\acmConference[CIDR '27]{Conference on Innovative Data Systems Research}{January 2027}{Amsterdam, The Netherlands}
\acmBooktitle{CIDR '27}
\acmDOI{}
\acmISBN{}

% placeholders for numbers pending the frozen campaign / ablations
\newcommand{\tbd}[1]{\textcolor{red}{\textbf{[#1]}}}

\begin{document}

\title{The Database as the Trusted Computing Base for LLM Agents:\\
Experience from TGMS}

\author{Xiaofei Zhang}
\affiliation{%
  \institution{University of Memphis}
  \city{Memphis}\state{TN}\country{USA}}
\email{xiaofei.zhang@memphis.edu}

\begin{abstract}
LLM agents are becoming a dominant client of data systems, and today's
interfaces ask them to do exactly what they are worst at: generate a
correct query in a general-purpose language, judge whether a partial
result is complete, perform arithmetic over it, and assert conclusions.
We argue for a different division of labor: \emph{the database should be
the trusted computing base for LLM agents}. An agent-native database
exposes a small, typed, costed operator algebra instead of unrestricted
query generation; executes plans deterministically over an explicit
belief state; and returns evidence-carrying results against which every
downstream claim can be mechanically checked.

We report experience from TGMS, a bi-temporal graph management system
built on this architecture and exercised with live open-source models on
temporal-graph analytics, a setting where computation is compositional,
historical corrections matter, and completeness cannot be judged from a
result page. Three lessons generalize beyond graphs: input schemas are
half a contract --- models invent \emph{output} fields, and statically
checking result paths took execution success from 0.00 to 1.00; evidence
support is multidimensional --- a model can compute a correct count over
a truncated page, so verification must track completeness, belief time,
and provenance, not just values; and agent accountability requires
belief-state semantics --- ``what did we believe on March~1'' is
unanswerable after a correction unless transaction time is a first-class
axis. A verifier built on these ideas rejects
500/500 injected numeric and identifier faults and, across
database-specific mutation classes (wrong belief state, truncated
completeness, uncited entities), detects
\tbd{per-class table, receipted run} with zero false positives, while
the architecture keeps prompts constant-size as data grows.
\end{abstract}

\maketitle

\section{The problem and the position}
\label{sec:position}

Consider an investigation question posed to an agent over a
communication network:
\begin{quote}\itshape
Under what we believed on March~1 --- before the record was corrected
--- among the accounts temporally reachable from Alice during February,
how many participated in a communication triangle that completed within
24 hours?
\end{quote}
Answering it requires resolving \emph{Alice} to an identifier, selecting
a historical belief state, time-respecting reachability, temporal motif
counting, exact arithmetic, and evidence that no intermediate result was
silently truncated. Every mainstream agent--database interface fails
this question in a characteristic way. Text-to-SQL/Cypher asks the model
to emit one correct query in a language that cannot express ``what we
believed on March~1'' over a latest-state store, and whose output the
model must still interpret. Retrieval pipelines serialize edges into
prompt chunks, destroying the structure the question quantifies over ---
and on numeric-dense data even $k{=}2$ retrieved chunks overflow a 28k
serving window (we measured 36{,}956 prompt tokens), so the model sees
under 1\% of the corpus. And in all of these designs the model performs
arithmetic, invents identifiers, and asserts numbers that appear in no
result.

The community is converging on the diagnosis: agents are becoming the
dominant database client, and systems should be redesigned for
them~\cite{liu2026agentfirst,urban2024caesura,wang2025aop}. We advance a
specific architectural answer:

\begin{quote}
\textbf{Position.} The database should be the \emph{trusted computing
base} (TCB) for LLM agents. The model plans and narrates; everything
that must be true --- computation, temporal semantics, evidence,
verification --- happens inside the database boundary, under contracts
the model cannot violate.
\end{quote}

Three design principles follow, each backed by a mechanism and a
measurement in this paper:

\textbf{P1 --- Agents need a contract-bearing algebra, not query
generation.} The agent selects and composes operations; it never
controls arbitrary computation. Every operation declares typed inputs
\emph{and outputs}, cost and completeness behavior, temporal semantics,
and provenance identity (\S\ref{sec:arch}).

\textbf{P2 --- Evidence support is multidimensional.} A result is not
just a value; it is a value under conditions:
$\langle$correctness, completeness, belief time, approximation,
provenance$\rangle$. Databases should return those conditions, and
verifiers should refuse full support when any dimension is degraded
(\S\ref{sec:evidence}).

\textbf{P3 --- Agent answers require a belief state, not a data
snapshot.} Accountability questions --- what was known when, which
correction invalidated it, can a decision be reproduced --- need
transaction time as a first-class axis (\S\ref{sec:evidence}).

TGMS~\cite{tgms2026} is our concrete instance: a bi-temporal property
graph store whose only query surface is thirteen verified temporal
operators exposed as agent tools, with a static plan verifier, a
deterministic executor producing content-addressed traces, and a claim
verifier that gates final answers against those traces. We use temporal
graph analytics as the proving ground precisely because it is
unforgiving: computation is compositional, corrections are routine,
completeness matters, and neither identifiers nor time can be left to
the model. The lessons, however, are about databases, not graphs.

\section{An agent-native architecture}
\label{sec:arch}

\begin{figure}[t]
\centering
\resizebox{\columnwidth}{!}{%
\begin{tikzpicture}[
  every node/.style={font=\small},
  box/.style={rounded corners=4pt, align=center, inner ysep=5pt,
              inner xsep=4pt, draw, thick},
  llm/.style={box, draw=orange!80!black, fill=orange!8},
  tcb/.style={box, draw=blue!60!black, fill=blue!6},
  arr/.style={-{Stealth[length=2.4mm]}, thick, black!60}]
\node[llm, text width=21mm] (plan) at (0,0) {\textbf{Planner}\\[-1pt]
  {\scriptsize emits plan IR}};
\node[llm, text width=21mm] (rep) at (0,-3.1) {\textbf{Reporter}\\[-1pt]
  {\scriptsize writes claims}};
\node[tcb, text width=24mm] (ver) at (3.6,0) {\textbf{Static verifier}\\[-1pt]
  {\scriptsize types $\cdot$ grounding $\cdot$\\[-2pt] output fields $\cdot$ cost}};
\node[tcb, text width=24mm] (exec) at (7.2,0) {\textbf{Executor}\\[-1pt]
  {\scriptsize deterministic $\cdot$\\[-2pt] content-addressed}};
\node[tcb, text width=24mm] (store) at (7.2,-1.6)
  {\textbf{Bi-temporal store}\\[-1pt]
  {\scriptsize valid $\times$ transaction time}};
\node[tcb, text width=24mm] (cver) at (3.6,-3.1) {\textbf{Claim verifier}\\[-1pt]
  {\scriptsize re-checks $\cdot$ gates}};
\node[draw=blue!60!black, dashed, thick, rounded corners=6pt,
      fit=(ver)(exec)(store)(cver), inner sep=7pt,
      label={[blue!60!black,font=\scriptsize\bfseries]below right:{TRUSTED
      (deterministic, contract-enforced)}}] {};
\node[font=\scriptsize\bfseries, text=orange!80!black] at (0,-1.55)
  {UNTRUSTED (LLM)};
\draw[arr] (plan) -- node[above,font=\scriptsize]{plan} (ver);
\draw[arr] (ver) -- node[above,font=\scriptsize]{valid} (exec);
\draw[arr, dashed] (ver.north) to[bend right=14]
  node[above,font=\scriptsize]{repair payloads} (plan.north);
\draw[arr] (exec) -- (store);
\draw[arr] (exec.south) to[bend left=12]
  node[right,font=\scriptsize,pos=0.35]{trace} (cver.east);
\draw[arr] (rep) -- node[above,font=\scriptsize]{answer object} (cver);
\draw[arr, dashed] (cver.west) to[bend left=18]
  node[left,font=\scriptsize]{summaries} (rep.west);
\end{tikzpicture}}
\caption{The trust boundary, not the module list. The LLM only plans and
narrates; plans are checked before execution, computation and temporal
semantics live inside the boundary, and claims are re-checked against
the trace before an answer leaves the system.}
\label{fig:tcb}
\end{figure}

Figure~\ref{fig:tcb} shows the architecture as a trust boundary. The
model appears exactly twice, both times outside the boundary; nothing it
produces is executed or emitted unchecked.

\paragraph{Plan IR, not query text.}
The agent expresses intent as a small JSON DAG. Each step names one
operator from a closed registry; references between steps use a
deliberately weak path language (dotted fields, row projections) that
cannot express computation. An \texttt{answer\_spec} names the step and
field that constitute the result. This is the CAESURA
intuition~\cite{urban2024caesura} pushed to its safety conclusion: if
the plan is data, the system can check it.

\paragraph{Contracts on both sides of every operator.}
Each operator declares its argument schema \emph{and its output fields},
cost estimation, pagination behavior, and temporal parameters. The same
registry generates the tool manual the model reads and the checks the
verifier enforces, so documentation and enforcement cannot drift. A
static verifier rejects, before execution: schema violations, cycles,
out-of-scope references, references to output fields the producing
operator does not emit, identifiers that neither the task supplied nor a
resolution operator produced (the \emph{grounding rule}), temporal
nonsense, and plans whose estimated cost exceeds a budget --- returning
structured repair payloads that name the violation and the legal
alternatives. Arithmetic is an operator too: the model never adds.

\paragraph{Deterministic, evidence-carrying execution.}
The executor runs the DAG with canonical serialization; every step
yields a content-addressed \texttt{result\_digest} (SHA-256), and
re-execution over the same store state is byte-identical. Results are pages with explicit
\texttt{rows\_total} and truncation flags, and \emph{truncation taint}
propagates to every dependent step. The reporter model turns trace
summaries into a typed answer object in which every numeric, entity,
ordering, and pattern claim cites the trace steps that ground it; the
claim verifier re-checks each claim against the cited digests and gates
unsupported claims out of the emitted answer.

This division reframes the question asked by recent work on LLM query
planning~\cite{urban2024caesura,wang2025aop} and agent-first
systems~\cite{liu2026agentfirst}: not \emph{can} a model produce a plan,
but what interface makes a fallible plan \emph{safe} --- checkable
before execution, auditable after.

\section{Evidence and belief semantics}
\label{sec:evidence}

\paragraph{Corrections are not updates.}
On March~1 the store believes fact $F$; on March~10 it learns $F$ was
wrong. A latest-state system overwrites and the question ``what did we
believe on March~1'' becomes unrepresentable --- the information was
destroyed at write time. TGMS stores half-open valid-time and
transaction-time intervals per version~\cite{snodgrass1999} and
distinguishes \emph{evolution} (the world changed) from \emph{correction}
(we were wrong). Every operator takes \texttt{as\_of\_tt}; a plan pinned
to a past belief state is reproducible byte-for-byte even after later
corrections (a tested metamorphic property). Derived artifacts join in:
summaries record the transaction time they were computed at, and a
correction quarantines every summary whose window it touches. This is
what agent accountability needs and agent-memory systems are groping
toward~\cite{zep2025,wang2026toki}: not temporal storage for its own
sake, but the ability to say \emph{what was known when, which correction
invalidated it, and whether an earlier decision reproduces under the
earlier belief state}.

\paragraph{Support is a tuple, not a boolean.}
In a live run, our 14B model called reachability, received the default
page of 100 rows (of 343), and computed a flawlessly correct count ---
of the page. A value-only verifier marks that claim supported; it is
false as a statement about the database. We now treat support as
\[
\langle \text{correctness},\ \text{completeness},\ \text{belief time},\
\text{approximation},\ \text{provenance} \rangle,
\]
and a claim whose evidence is incomplete on any dimension can be at most
\emph{weakly supported}. The mechanism is mundane --- flags on result
pages, taint through the DAG, verdict caps in the verifier --- and that
is the point: it generalizes immediately to SQL \texttt{LIMIT}, top-$k$
retrieval, sampling, approximate query processing, timeouts, stale
replicas, and partially failed distributed queries. Databases already
know when their answers are partial; they just don't say so in a form an
agent pipeline is forced to respect.

\section{The TGMS prototype and lessons}
\label{sec:lessons}

TGMS implements the architecture in ${\sim}10$k lines of Python over
embedded columnar and graph backends: an append-only event log as the
source of truth, thirteen operators in five groups (belief-state reads;
snapshots and diffs; time-respecting traversal; temporal aggregation and
patterns; resolution and compute), each validated against an independent
brute-force oracle on 500 randomized cases, plus the planner/verifier
loop served by vLLM~\cite{kwon2023vllm} with open-weight
models~\cite{qwen25} on a single 24\,GB GPU. Three lessons from live
traffic did not appear in any design document.

\paragraph{Lesson 1: models invent \emph{output} fields.}
Our first live matrix run scored 0.00 execution success on every task
--- with every plan passing validation. Plans referenced
\texttt{s2.count} where the operator emits \texttt{rows\_total}: input
schemas constrain what goes into a tool, and nothing constrained what
the model read back. Declaring output fields in the registry and
checking every reference statically took probe-task execution success
from 0.00 to 1.00, and structured repair payloads (``available:
\texttt{rows, rows\_total, truncated}'') roughly tripled overall
execution success at 7B. Result schemas deserve the same rigor as
argument schemas; tool manuals must be checkable, not just readable.

\paragraph{Lesson 2: constrained decoding is not the safety
mechanism.} Grammar-constrained JSON decoding --- the standard remedy
for malformed plans --- made everything worse at 14B: exact match
0.409${\to}$0.227, first-emission validity 0.50${\to}$0.32, and 65\%
more wall time. Syntactic validity was never the bottleneck; semantic
plan quality was, and grammar masks degrade it. Checking artifacts
\emph{after generation} against contracts, with a repair loop, is both
safer and cheaper than constraining generation itself.

\paragraph{Lesson 3: the verifier's blind spots are database
semantics.} Value-grounding catches fabrication but not mis-citation:
small counts coincidentally appear in other steps' payloads, so a claim
citing the wrong step often still ``grounds.'' We now reject claims
whose provenance pointer names an uncited step. Conversely, entity
\emph{under}-claiming is invisible to trace grounding by construction
--- precision is checkable against evidence; completeness of a reported
set is checkable only against the database. Verification design is
database design.

\section{Preliminary evidence}
\label{sec:eval}

All numbers carry receipts (git SHA, config SHA, frozen-split SHA) and
regenerate from the public artifact~\cite{tgms2026}. Tasks are
program-generated with engine-computed gold over the CollegeMsg
network~\cite{panzarasa2009}, the email-EU log, and a synthetic
temporal-pattern workload; test splits (94/94/102 tasks) were frozen,
with SHAs logged, before any test measurement.

\paragraph{End-to-end.}
Table~\ref{tab:e2e} reports the frozen-test campaign
\tbd{final numbers landing; dev-split results shown here are complete}.
On the CollegeMsg dev split (22 tasks, 14B), the operator-backed agent
reaches 0.409 exact match versus 0.045--0.182 for vector-RAG,
static-graph RAG, and text-to-Cypher under identical serving; on
correction probes it scores 0.67 while all baselines score zero at 14B
--- not a capability gap but a representational one: the baselines'
inputs no longer contain the earlier belief state. The emitted-answer
unsupported-claim rate is 0.000 with gating; the no-verifier ablation's
raw answers carry \tbd{UCR, campaign} unsupported claims.

\begin{table}[t]
\centering\small
\caption{Frozen-test campaign (placeholder layout): exact match by task
family, 5 systems $\times$ 3 seeds, Qwen2.5-14B; second family
\tbd{Phi-4-mini} for architecture portability.}
\label{tab:e2e}
\begin{tabular}{lcccc}
\toprule
 & TGMS & Vec-RAG & Graph-RAG & Text-to-Cypher \\
\midrule
CollegeMsg (94) & \tbd{x} & \tbd{x} & \tbd{x} & \tbd{x} \\
email-EU (94)   & \tbd{x} & --      & \tbd{x} & \tbd{x} \\
synthetic (102) & \tbd{x} & --      & \tbd{x} & \tbd{x} \\
correction probes & \tbd{x} & \tbd{0.00} & \tbd{0.00} & \tbd{0.00} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Mechanism ablations.}
Removing output contracts \tbd{E1: invalid references, repairs,
execution success, accuracy}. Removing completeness propagation: in a
controlled generator that shrinks one page limit so a correct count runs
over a truncated page, the verifier refuses full support in 15/15 cases
with taint on and \emph{misses 15/15} with taint off --- the mechanism
is load-bearing, not decorative \tbd{E2 end-to-end rates on the frozen
suites}. Removing transaction time is not an ablation but an
impossibility: the latest-state representation does not contain the
queried information, which is why every baseline scores zero on probes.

\paragraph{Verifier validation.}
Beyond the classic classes (500/500 injected count and identifier faults
detected; 0 false positives on 87 clean answers), database-specific
mutation classes on a dedicated suite (202 enriched clean answers, 0
false positives): real-but-uncited entity added 100/100 detected;
ordering operands swapped 100/100; microsecond/millisecond unit
confusion 100/100; \emph{correct value from the wrong belief state}
60/60; wrong-step citation 36/100 (the remainder are genuinely grounded
by the surrogate step); entity member dropped 0/100 --- reported, not
hidden: grounding checks precision, and set completeness is checkable
only against the database (Lesson~3)
\tbd{receipted rerun on frozen-campaign hardware}.

\paragraph{The database stays a database.}
Prompt size is constant (${\sim}$8--10k tokens) as data grows, because
structure never enters the context window. Operator p50 latency on
$10^5$/$10^6$/$10^7$-event stores: 2-hop snapshot 13/99/272\,ms; global
diff 23/163/485\,ms; metric time series 144/157/1127\,ms (40-thread CPU
host, DuckDB backend; the $10^7$ store is 2.6\,GB). Traversals beyond
the cost budget refuse with a narrowing suggestion rather than degrade
--- bounded execution is part of the contract
\tbd{small line figure replacing this prose}.

\paragraph{Honest limitations.}
Single-GPU open-weight serving; one graph domain; 290 frozen test tasks;
pattern claims are verified but not yet gated; agent write-back is
schema-ready but disabled pending provenance policy. Sustained agent
traffic also surfaced a serving lesson: on older accelerators the
inference engine degrades and must be recycled on a schedule ---
agent-era benchmarking is an endurance discipline.

\section{Research agenda}
\label{sec:agenda}

TGMS is one point in a space the community should map:
\emph{agent-facing logical algebras} --- what is the right closed
operator set for relational, document, and streaming data, and can it be
derived from a workload?; \emph{cost-based rewriting of agent plans} ---
the verifier sees the whole DAG before execution, so optimize it,
don't just admit it; \emph{evidence-aware execution} --- make
completeness, staleness, and approximation first-class result
attributes with propagation semantics; \emph{verification under
approximation} --- what does ``supported'' mean over samples and
sketches?; \emph{bi-temporal provenance} --- derived-view and
memory invalidation under correction, at scale; and \emph{multi-agent
isolation} --- belief states as the unit of concurrency control when
many agents read, and eventually write, one store.

\medskip
\noindent Today's database interfaces assume a competent client. LLM
agents violate that assumption. Agent-native databases must therefore
make plans checkable, computations deterministic, evidence explicit, and
conclusions auditable. TGMS is an initial architecture showing what such
a database can look like.

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
