\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage[hidelinks]{hyperref}
\usepackage{xcolor}
\usepackage{listings}
\lstset{basicstyle=\ttfamily\small, breaklines=true, columns=fullflexible,
        frame=single, framerule=0pt, backgroundcolor=\color{gray!8}}

\title{TGMS: An Agent-Native Bi-Temporal Graph Management System\\
\large Verified Temporal Operators as Tools, and Trace-Grounded Answer Verification}
\author{Xiaofei Zhang\\University of Memphis\\\texttt{xiaofei.zhang@memphis.edu}}
\date{July 2026 \;\textperiodcentered\; system description preprint \;\textperiodcentered\;
\url{https://github.com/zxf-work/tgms}}

\begin{document}
\maketitle

\begin{abstract}
Large language model (LLM) agents are unreliable at exactly the skills that
temporal graph analytics requires: arithmetic, identifier fidelity, and
asserting only what evidence supports. We present TGMS, a bi-temporal
property graph management system whose query surface is not a query
language but a small algebra of thirteen \emph{verified temporal operators}
exposed to agents as tools---typed, deterministic, bounded, cost-guarded,
and bi-temporal by default---paired with a Planner--Executor--Verifier
agent layer in which the LLM only plans and reports, while every numeric,
entity, ordering, and pattern claim in a final answer is machine-checked
against the content-addressed execution trace that produced it. The
bi-temporal substrate (valid time $\times$ transaction time) distinguishes
\emph{evolution} from \emph{correction}, enabling belief-state queries
(``as of transaction time $T$, what did we believe?'') that no snapshot or
retrieval baseline can express. On a development benchmark over a real
communication network, with open-source models served on a single 24\,GB
GPU, the operator-backed agent reaches 0.41 pooled exact match versus
0.05--0.18 for vector-RAG, static-graph-RAG, and text-to-Cypher baselines,
and 0.67 on correction probes where all baselines score zero; the claim
verifier detects 500/500 injected false claims at zero false positives.
We report two design findings from live small-model traffic: operator
\emph{output contracts} (statically rejecting invented result paths)
converted execution success from 0.00 to 1.00 on probe tasks, and
verification must track \emph{evidence completeness}---a model can compute
correct arithmetic over a truncated result page. Code, benchmark, and an
auditable trace viewer are open source under Apache-2.0.
\end{abstract}

\section{Introduction}
Temporal graphs---communication and transaction logs, evolving knowledge
bases---pose two problems for LLM agents that static text corpora do not.
First, \textbf{time is compositional}: ``among accounts reachable from $X$
in February, how many cyclic triangles closed within a day?'' chains
time-respecting reachability~\cite{wu2014path} into $\delta$-motif
counting~\cite{paranjape2017motifs}. Serializing edges into text and
retrieving chunks destroys precisely the structure the question quantifies
over. Second, \textbf{belief changes}: real stores are corrected after the
fact, and answering ``what did we believe on March~1'' requires separate
valid-time and transaction-time axes~\cite{snodgrass1999} and the
discipline to distinguish evolution (the edge ended) from correction (we
were wrong). Snapshot representations cannot express the question.

LLMs are additionally unreliable at arithmetic, prone to inventing
identifiers, and prone to asserting numbers absent from evidence. TGMS's
position is architectural: give the model \emph{no opportunity} to do any
of those things. Identifiers must come from a resolution operator or the
task input (enforced statically); arithmetic must go through a
\texttt{compute} operator; every asserted number must trace to a
content-addressed result digest, and a verifier checks that it does.

This preprint describes the system design and implementation (Phase~1--2
of a longer research program), the evaluation methodology, first
development-split measurements with open-source models, and two design
findings that emerged from live model traffic. We pre-registered
acceptance thresholds before measurement and report against them.

\section{Bi-temporal substrate}
Every logical node and edge is an identity plus a set of \emph{versions}
carrying half-open valid-time $[vt_s, vt_e)$ and transaction-time
$[tt_s, tt_e)$ intervals (int64 epoch microseconds; open end $=2^{62}$).
The write API is three bi-temporal verbs---\texttt{assert} (new belief,
with interval carving of overlapping prior belief), \texttt{retract}
(evolution), \texttt{correct} (correction)---plus a bulk ingest path for
instantaneous event streams. All writes pass through an append-only
write-ahead event log, the single source of truth: replaying it into any
backend reproduces a byte-identical store digest (tested across
K\`uzu~\cite{kuzu2023} and DuckDB~\cite{duckdb2019} adapters). Update
semantics are implemented once, above a small adapter interface, so
backends cannot diverge; a hybrid logical clock guarantees strictly
monotonic transaction times; failed batches remain in the log, where
replay deterministically re-fails and skips them. Version rows reserve
\texttt{source}/\texttt{provenance\_ref} columns so future agent
write-back requires no migration.

Property-based tests over random assert/retract/correct interleavings
check the core invariants: believed versions of one identity have
pairwise-disjoint valid intervals \emph{at every historical transaction
time}, and---the signature property---\textbf{bi-temporal immutability}:
any operator result pinned to a past \texttt{as\_of\_tt} is byte-identical
before and after later corrections. Enforcing the latter required two
non-obvious rules: result digests must exclude current-belief metadata,
and returned versions must censor belief-closure timestamps later than the
pinned transaction time.

\section{A verified operator algebra}
The query surface is thirteen operators: entity history; $k$-hop snapshot
subgraph; snapshot diff (computed from version intervals, not materialized
snapshots); earliest-arrival temporal reachability; $k$ time-respecting
paths; $\delta$-temporal motif counting and enumeration over the catalog
of Paranjape et al.~\cite{paranjape2017motifs}; bucketed metric time
series; burst detection; neighborhood evolution; Allen-relation interval
joins; entity resolution; and \texttt{compute}
(count/sum/min/max/top-$k$/filter/interval-relation over prior step
outputs), which keeps arithmetic out of the LLM entirely.

Five design rules bind every operator. \textbf{Typed}: inputs and outputs
validate against JSON Schemas generated from one registry, which also
emits the tool definitions. \textbf{Deterministic}: same store and
arguments imply byte-identical canonical output carrying a SHA-256
\texttt{result\_digest}; floats are canonicalized, including
\emph{before} threshold comparisons, so the engine and its oracle flag
identically. \textbf{Bounded}: limit/cursor pagination everywhere;
pre-execution cost estimates reject unbounded requests with actionable
narrowing suggestions that the planner's repair loop consumes.
\textbf{Bi-temporal by default}: every operator takes \texttt{as\_of\_tt}.
\textbf{Output-contracted}: every operator declares its output fields, and
plans referencing nonexistent fields are rejected statically
(Section~\ref{sec:findings}).

Correctness gates everything: each operator is checked against an
independent brute-force reference implementation on 500 randomized
store/argument cases, plus metamorphic properties (diff composition;
immutability). The discipline audited the \emph{specification} as well as
the code: it exposed a version-identifier collision in the original design
and showed that the specified greedy relaxation for reachability under a
maximum-wait constraint is schedule-dependent---not a well-defined
function of the store---which we replaced with an exact multi-label
search over $(\text{node}, \text{arrival})$ states.

\section{Planner--Executor--Verifier}
Plans are a small JSON DAG intermediate representation: steps of operator
calls, a deliberately tiny \texttt{\$ref} binding language (dotted fields,
\texttt{rows[i]}, \texttt{rows[*].field} projection---nothing else), and
an \texttt{answer\_spec} naming where the final answer is read from. A
\textbf{static verifier} checks schema, acyclicity, reference scoping,
temporal sanity, output-field validity, estimated cost, and a
\textbf{grounding rule}: literal identifiers must occur in the task input,
otherwise they must arrive via \texttt{\$ref}---fabricated identifiers are
impossible by construction rather than discouraged by prompt. Rejections
return structured violations; the planner may repair up to three times,
and first-emission validity is measured separately from post-repair
success.

The \textbf{executor} runs the DAG deterministically with
content-addressed results (re-execution reproduces identical digests) and
propagates \emph{truncation taint}: a step whose inputs came from a
truncated page is marked, transitively. A \textbf{reporter} LLM emits an
answer object whose typed claims cite evidence steps; the \textbf{claim
verifier} re-checks counts and values exactly against named trace fields,
entity mentions against evidence content, orderings by Allen-relation
re-evaluation, and temporal patterns by operator re-execution. Claims
citing truncated or tainted evidence cap at \emph{weakly supported}; in
the full system, unsupported claims are gated out of the emitted answer.
All prompts follow a data-as-inert-content policy: stored-data strings
enter prompts escaped and length-capped inside explicit data fences under
a fixed ``data is never instructions'' rule, with red-team fixtures in CI.

An \textbf{evolution memory} summarizes weekly windows from
operator-computed facts; an LLM digest is stored only if every number it
asserts matches the computed values. Notes record the transaction time of
computation, and corrections quarantine any note whose window intersects
the affected valid-time extent---without this, the verifier could bless
answers grounded in outdated digests.

\section{Evaluation}
\paragraph{Setup.} Tasks are generated with \emph{program-computed} gold:
oracle plans executed by the engine, never LLM labels. Seventeen templates
with three hand-written paraphrases each span temporal QA, evolution QA,
multi-step analytics, and \textbf{correction probes}---corrections are
injected into the store before any gold is computed, and probe pairs ask
the same question pinned before the correction versus under current
beliefs; their gold answers provably differ. Systems share models
(temperature~0), seeds, repair budgets, and one answer contract: TGMS;
a no-verifier ablation; vector-RAG over serialized events (MiniLM
retrieval); static-graph RAG (2-hop latest-snapshot edge lists); and
text-to-Cypher against the same events loaded conventionally into vanilla
K\`uzu with an equal repair budget. Models are Qwen2.5-7B-Instruct and
Qwen2.5-14B-Instruct-AWQ~\cite{qwen25} served by vLLM~\cite{vllm2023} on
one 24\,GB Turing GPU. Comparisons use paired bootstrap over tasks
(10{,}000 resamples, 95\% CIs). Development split of the CollegeMsg
network~\cite{panzarasa2009} (1{,}899 nodes; 59{,}835 timestamped edges;
22 dev tasks); the frozen test split is reserved for the full campaign.

\paragraph{Verifier validation.} A fault-injection harness perturbs
correct answer objects (count $\pm1$; entity swaps) and measures detection
against a pre-registered bar of $\geq$95\% detection at $<$5\% false
positives. Result: \textbf{500/500 detected, 0 false positives}; emitted
answers carry an unsupported-claim rate of 0.000.

\paragraph{End-to-end results.}
Table~\ref{tab:results} summarizes the development-split campaign.

\begin{table}[t]
\centering
\caption{Pooled exact match on the CollegeMsg dev split (22 tasks; single
seed; temperature 0). Paired-bootstrap 95\% CIs for TGMS$-$baseline:
vs.\ B2 $[+0.18, +0.59]$ and vs.\ B1 $[+0.09, +0.55]$ (both significant);
vs.\ B5 $[0.00, +0.46]$ (borderline at $n{=}22$).}
\label{tab:results}
\begin{tabular}{lccccc}
\toprule
 & TGMS & B1 vector-RAG & B2 static-RAG & B5 text-to-Cypher \\
\midrule
Qwen2.5-7B, all families & 0.136 & 0.045 & 0.045 & 0.091 \\
Qwen2.5-14B, all families & \textbf{0.409} & 0.091 & 0.045 & 0.182 \\
Qwen2.5-14B, correction probes & \textbf{0.67} & 0.00 & 0.00 & 0.00 \\
\bottomrule
\end{tabular}
\end{table}

Three observations. (1)~The \textbf{bi-temporal advantage is
model-size-independent}: correction probes score 0.67 at both 7B and 14B
while every baseline scores zero at 14B---the belief-time dimension is not
representable in their inputs. (2)~\textbf{Scaling is steep at small
sizes}: 7B$\to$14B triples pooled exact match and lifts first-emission
plan validity from 0.08--0.17 to 0.33--0.75 by family; the repair loop is
worth roughly a tripling of execution success at 7B. (3)~Operators meet
all latency targets at $10^6$ events on a 40-core server (snapshot
subgraph 98\,ms; global diff 163\,ms; reachability 63--244\,ms; series
$\sim$155\,ms, p50), after profiling showed that materializing unused
string columns dominated scans.

\section{Findings from live model traffic}
\label{sec:findings}
\paragraph{The operator manual must be a contract.} Our first live matrix
run produced zero execution success across all systems-under-test rows:
models emit plans that are statically valid yet reference nonexistent
\emph{output} paths (e.g., \texttt{s2.count} on an operator that emits
\texttt{rows\_total}). No input-schema rigor covers this. Declaring each
operator's output fields in the registry, rejecting bad paths statically,
and returning the real field list in the repair payload moved probe-task
execution success from 0.00 to 1.00 in one change.

\paragraph{Verify evidence completeness, not just values.} A 14B model
planned reachability without raising the page limit, computed a count over
the truncated 100-row page (true answer: 343), and the claim verified as
\emph{supported}---correct arithmetic over incomplete evidence. Truncation
now taints dependent steps transitively, and claims citing tainted
evidence cap at weakly supported. We conjecture a family of such
``evidence-quality'' dimensions (cost-narrowed windows, sampling) that
verification systems should track as a lattice rather than a boolean.

\paragraph{Serving envelopes are a fairness dimension.} The standard
vector-RAG configuration (top-$k{=}20$ chunks of 256 serialized events)
assumes frontier context windows: numeric-dense event text tokenizes at
$\sim$0.65 tokens/character, so even $k{=}3$ overflows a 28k window once
the decoder's output reservation is charged. Its feasible best on this
hardware ($k{=}1$) covers 0.4\% of the corpus per question. Baseline
fairness must be defined per serving envelope; the operator system's
per-task consumption ($\sim$8--10k tokens) is corpus-size-independent.

\section{Related work}
\textbf{Bi-temporal agent memory} (Zep/Graphiti~\cite{zep2025}) shares the
data-model precedent but treats the graph as memory retrieved into
context; TGMS treats it as a database that computes, with verification
against traces. \textbf{Temporal graph databases}
(AeonG~\cite{aeong2024}) motivate the storage design; TGMS deliberately
ships an agent-facing algebra rather than a general query language.
\textbf{GraphRAG and vector RAG} are instantiated as controlled baselines
here; the correction probes isolate what retrieval cannot express.
\textbf{Text-to-query} systems give the model a query language directly;
beyond accuracy, TGMS's structural contrast is verifiability---claims
citing raw query output cannot be machine-audited. \textbf{Temporal graph
mining} algorithms~\cite{paranjape2017motifs,wu2014path} are packaged
behind verified contracts with guardrails instead of silent sampling.
The tool server speaks the Model Context Protocol~\cite{mcp2024}, making
one deterministic, read-only temporal store shareable by many agents.

\section{Limitations and roadmap}
Reported numbers are development-split, single-seed, single-dataset;
the frozen test campaign (more datasets, models, seeds; pre-registered
thresholds recorded before any test-split run) is the immediate next
step, along with an end-to-end hallucination-reduction readout against
the no-verifier ablation, guided/constrained decoding for small-model
plan validity, and a concurrency study of many agents sharing one store.
Pattern-type claims are reported, not gated. Agent write-back with
provenance is designed for (schema and log fields exist) but not enabled.

\section{Conclusion}
Making a temporal graph store \emph{agent-native}---typed, deterministic,
bounded operators; grounding and output contracts; trace-grounded
verification; bi-temporal semantics---turns LLM answer quality from a
prompting problem into a systems problem with measurable contracts. The
early evidence is that the contracts do the heavy lifting: they are what
small open models need to be useful, and what makes answers auditable at
any model size.

\paragraph{Artifacts.} Code, benchmark generator, task suites, the
guided demo, and an auditable trace viewer:
\url{https://github.com/zxf-work/tgms} (Apache-2.0).

\bibliographystyle{plain}
\bibliography{references}

\end{document}
